## Cyclist data

In this report we are designing marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would be buy membership, and how digital media could affect thir marketing tactics.

This report shows all the steps that has been taken.


Loading Packages
```{r loading packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

Setting up the work directory
```{r message=FALSE, warning=FALSE}
setwd("S:/Projects/Cyclist")
getwd()
```

Uploading datasets
```{r message=FALSE, warning=FALSE}
q3_2019 <- read_csv("S:/Projects/Cyclist/Data/Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("S:/Projects/Cyclist/Data/Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("S:/Projects/Cyclist/Data/Divvy_Trips_2020_Q1.csv")
```

Comparing column names of each of the file
```{r}
colnames(q4_2019)
colnames(q3_2019)
colnames(q1_2020)
```

Renaming cloumn name to make them consistent
```{r}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
```

Inspect the dataframes and look for inconsistency
```{r}
str(q1_2020)
str(q4_2019)
str(q3_2019)
```

Convert ride_id and rideable_type to character so that they can stack correctly
```{r}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

Stacking individual quarter's data frames into one big data frame
```{r}

all_trips <- bind_rows(q3_2019, q4_2019, q1_2020)

```

Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020
```{r}

all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))

```

### ==CLEANING DATA==
Inspect the new table that has been created
```{r}

colnames(all_trips) #List of column names
nrow(all_trips) #Number of rows in data frame
dim(all_trips)  #Dimensions of the data frame
head(all_trips)  #See the first 6 rows of data frame
str(all_trips) #See list of columns and data types
summary(all_trips) #Statistical summary of data.
```

seeing how many observations fall under each usertype
```{r}

table(all_trips$member_casual)
```

Reassign to the desired values (going with the current 2020 labels)
```{r}

all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

Check to make sure the proper number of observations were reassigned
```{r}

table(all_trips$member_casual)
```

Add columns that list the date, month, day, and year of each ride
This will allow us to aggregate ride data for each month, day, or year 
before completing these operations we could only aggregate at the ride level
```{r}

all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Add a "ride_length" calculation to all_trips (in seconds)
```{r}

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

Inspect the structure of the columns
```{r}

str(all_trips)
```

Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}

is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Remove "bad" data
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
We will create a new version of the dataframe (v2) since data is being removed
```{r}

all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

### ==CONDUCT DESCRIPTIVE ANALYSIS==

Descriptive analysis on ride_length (all figures in seconds)
```{r}

mean(all_trips_v2$ride_length) #straight average (total ride length / rides)

median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths

max(all_trips_v2$ride_length) #longest ride

min(all_trips_v2$ride_length) #shortest ride
```

Consolodating four lines into one
```{r}

summary(all_trips_v2$ride_length)
```

Compare members and casual users
```{r}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

average ride time by each day for members vs casual users
```{r}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Days of the week are out of order. Let's fix that.
```{r}

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

average ride time by each day for members vs casual users
```{r}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

analyze ridership data by type and weekday
```{r}

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)		
```

visualize the number of rides by rider type
```{r}

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

Let's create a visualization for average duration
```{r}

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

### ==EXPORT SUMMARY FILE==
```{r}

counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

write.csv(counts, file = 'S:/Projects/Cyclist/Data/avg_length_ride.csv')

```

